---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(forcats)
library(broom)
library(ggbeeswarm)
library(data.table)
library(reshape2)
library(ggridges)
```

Load datasets. Ignore everything except DLI patients. Check number of clonotypes in donors and receptients - we have engough clones for statistics everywhere. We group clones by their abundance in donors: 1 (singletons), 2 (doubletons), 3 (tripletons, as good as doubletons, no need to hate them Vanya) and 4+ reads (Large). The choice is dictated by observing the fact that for rare events Poisson distribution shows huge difference in capture probability for $\lambda \in [1,3]$ while smaller $\lambda$ values are unlikely to be encountered and quantified. Moreover, for large clones, each hyperexpanded variant has its own history and, likely, its own dynamic, so binning them to different bins based on minor differences in frequency (e.g. $0.1\%$ vs $0.01\%$) makes little sense.

```{r}
data <- list.files("original", full.names = T) %>%
  as.list %>%
  lapply(function(x) fread(x) %>% mutate(sample.id = x)) %>%
  rbindlist %>%
  mutate(sample.id.old = sample.id,
         dli = !str_starts(sample.id.old, fixed("original/sh.p")),
         sample.id = paste0("D", sample.id %>% as.factor %>% as.integer, ifelse(dli, "*", ""))) #%>%
  #filter(dli)

data %>%
  select(sample.id.old, dli, sample.id) %>%
  unique

data <- data %>%
  mutate(donor.quantile = case_when(
    is.na(cloneCount.don) ~ "Missing",
    cloneCount.don == 1 ~ "Singleton",
    cloneCount.don == 2 ~ "Doubleton",
    cloneCount.don == 3 ~ "Tripleton",
    T ~ "Large"
    ))

data %>%
  ggplot(aes(x = sample.id, fill = donor.quantile)) +
  geom_bar() +
  theme_classic() 

data %>%
  filter(donor.quantile != "Missing") %>%
  ggplot(aes(x = donor.quantile, 
             fill = sample.id)) +
  geom_bar(position = "dodge") +
  scale_y_log10() +
  geom_hline(yintercept = 100, linetype = "dotted") +
  geom_hline(yintercept = 1000, linetype = "dashed") +
  theme_classic()
```

We split our donor dataset into singletons, doubletons, tripletons and higher-order clonotypes. Each of these subsets contains enough clones to reliably estimate the probability of recapturing a clonotype from a given subset of donor clonotypes. Interestingly, the ration between recapturing probabilities of singletons, doubletons and tripletons is in line with exponential difference stemming from Poisson distribution.

```{r}
# summarize & estimate parameters of beta distribution
alpha.prior <- 1
beta.prior <- 1
data.s <- data %>%
  filter(donor.quantile != "Missing") %>%
  group_by(sample.id) %>%
  mutate(total.don = sum(cloneCount.don, na.rm = T),
         clones.don = length(unique(aaSeqCDR3.don)),
         total.rec = sum(cloneCount.rec, na.rm = T),
         clones.rec = length(unique(aaSeqCDR3.rec))) %>%
  group_by(dli, sample.id, donor.quantile, total.don, clones.don, total.rec, clones.rec) %>%
  summarize(alpha = sum(!is.na(cloneCount.rec)) + alpha.prior,
            beta = sum(is.na(cloneCount.rec)) + beta.prior) %>%
  ungroup

data.sp <- data.s %>%
  merge(tibble(p = c(0:1000/1000, 10^(-4000:-1000/1000)))) %>%
  group_by(sample.id, donor.quantile) %>%
  mutate(Pbeta = dbeta(p, alpha, beta)) %>%
  ungroup

data.sp %>%
  group_by(sample.id) %>%
  mutate(height = Pbeta / max(Pbeta)) %>%
  ggplot(aes(x = p, y = sample.id, height = height, 
             fill = factor(donor.quantile, levels = c("Singleton",
                                                       "Doubleton",
                                                       "Tripleton",
                                                       "Large"))
             )) +
  geom_ridgeline(color = NA, alpha = 0.9) + 
  scale_x_log10("Capture probability", limits = c(0.8e-3, 1e-1)) + ylab("Pbeta") +
  scale_fill_brewer("", palette = "Spectral") +
  #facet_wrap(~sample.id, scales = "free_y") +
  theme_classic() +
  theme(aspect = 1, legend.position = "bottom")
```

Interestingly, the TCR recovery rate is related both to the total number of clones in donor and recipient (* - VANYA write an explainaton based on what I've told you yesterdat)

```{r}
data.s %>%
  ggplot(aes(x = clones.rec / clones.don, y = alpha / (alpha + beta), 
             color = factor(donor.quantile, levels = c("Singleton",
                                                       "Doubleton",
                                                       "Tripleton",
                                                       "Large")))) +
  geom_smooth(method = "lm", aes(), size = 1) +
  geom_point(size = 5) +
  geom_text(aes(label = sample.id), size = 2, color = "white") +
  scale_x_log10("Clones in receptient / clones in donor") + 
  scale_y_log10("Capture probability") +
  scale_color_brewer("", palette = "Spectral") +
  theme_classic() +
  theme(aspect = 1, legend.position = "bottom")
```

Quantifying the effect of various factors -- number of clones detected in donor, number of clones detected in receptient and the frequency quantile of a given clonotype in donor -- on the recapture probability. Log-transformed variables show extremely high correlation.

```{r}
data.coord <- data.s %>%
  group_by(donor.quantile) %>%
  mutate(logRecaptureProb = log(alpha / (alpha + beta)), 
         logClonesRecepient = log(clones.rec),
         logClonesDonor = log(clones.don)) %>%
  ungroup %>%
  mutate(donor.quantile = factor(donor.quantile, levels = c("Singleton",
                                                       "Doubleton",
                                                       "Tripleton",
                                                       "Large")))

data.lm <- data.coord %>%
  ungroup %>%
  mutate(donor.quantile = as.factor(donor.quantile)) %>%
  do(lm(.$logRecaptureProb ~ .$donor.quantile + .$logClonesRecepient + .$logClonesDonor) %>% tidy)

data.lm
data.coord %>%
  ungroup %>%
  mutate(donor.quantile = as.factor(donor.quantile)) %>%
  do(lm(.$logRecaptureProb ~ .$donor.quantile + .$logClonesRecepient + .$logClonesDonor) %>% aov %>% tidy) %>%
  mutate(var = sumsq / sum(sumsq) *100)
```

Origin of clones found in recepient: number of highly expanded clones that originated from expanded donor clones and rare donor clones varies and depends on donor.

```{r}
data %>%
  filter(!is.na(cloneCount.rec)) %>%
  group_by(sample.id) %>%
  mutate(rank = rank(-cloneCount.rec, ties.method	= "first"),
         freq.rec = cloneCount.rec / sum(cloneCount.rec)) %>%
  filter(donor.quantile != "Missing") %>%
  ggplot(aes(x = donor.quantile, y = rank)) +
  geom_hline(yintercept = 100, linetype = "dashed") +
  geom_quasirandom(aes(size = freq.rec, color = donor.quantile)) +
  geom_boxplot(fill = NA, color = "black", outlier.colour = NA) +
  coord_flip() +
  scale_y_log10("Clonotype rank") +
  xlab("") +
  scale_size_continuous("Clonotype size") +
  scale_color_brewer("", palette = "Spectral") +
  facet_wrap(~sample.id, scales = "free_x") +
  theme_classic() +
  theme(aspect = 1, legend.position = "bottom")
```

```{r}
```